\mytitle

### 2015-08-02 MSc_diss.tex
Finished section about ANs chemistry

\subsection{Alkyl nitrates}
Alkyl nitrates ($RONO_2$) are organic tropospheric trace gases, which through one of their formation mechanisms are closely linked with the production of tropospheric ozone \citep{Reeves2007}. This mechanism is present in the following section.

\subsubsection*{Alkyl nitrates chemistry}

The photochemical formation of the alkyl nitrates starts when the chemistry of the troposphere reaches the regime of $O_3$ production, i.e. when reactions of the peroxy radicals with $NO$ start to compete with their self- and cross-reactions more effectively due to an increase in $NO_x$ concentration. In this case the reaction between the alkyl peroxy radical ($RO_2$, where $R$ denotes methyl group, $CH_3$) and $NO$ can reveal its two-channel nature \citep{Day2003}, which is controlled by the branching ratio $\alpha_3$ in the following way \citep{Roberts1990}
\begin{subequations} \label{reac:RONO2ab}
\begin{align}
RO_2 + NO &\xrightarrow{1-\alpha_3} RO + NO_2 \label{eq:peroxy_no1}\\
RO_2 + NO &\xrightarrow{\alpha_3} RONO_2 \label{eq:peroxy_no2}
\end{align}
\end{subequations}
The latter reaction can serve as a termination step of the propagated $O_3$ forming cycle, and as \citep{Roberts1998} estimated this likely begin to occur when the $NO$ concentration increases up to 100-200 ppt.

\subsubsection*{Branching ratio}
The branching ratio $\alpha_3$ is larger when the parent alkane is larger and is a function of temperature and pressure.
\citep{Newland2013}
[The yield of this secondary pathway increases with increasing carbon chain length (Table 1.3)]

\subsubsection*{Yield}
\citep{Newland2013}
[Since the yield of alkyl nitrates from Reaction 1.9 increases with chain length but the mixing ratios of the alkanes tends to decrease with chain length, propyl and butyl nitrates are generally present at the highest concentrations in polluted air (Beyersdorf et al., 2010)]
\citep{Sommariva2008}
[The yield of alkyl nitrate formation is temperature and pressure dependent and increases with the number of carbons in the peroxy radical \citep{Roberts1990}!]

\subsubsection*{Alternative photochemical chemical sources}
\citep{Newland2013}
[Another photochemical pathway has been shown to become important for the production of the lower (methyl) alkyl nitrates in polluted air. Reaction 1.10 shows the reaction of the alkoxy radical, formed in in Reaction 1.9a, with NO2.
\begin{equation}\label{reac:RO+NO2=RONO2}
RO + NO_2 \rightarrow RONO_2
\end{equation}
This photochemical pathway has been shown to become dominant over Reaction 1.9b for the production of methyl nitrate in polluted air masses where NOx mixing ratios exceed 40 ppb (Archibald et al., 2007). This reaction is only likely to be important for methyl nitrate (!) rather than the higher nitrates because of the very small branching ratio ($\alpha_3$) for the production of methyl nitrate via the reaction of the peroxy radical with NO]
\citep{Newland2013}
[in highly polluted environments, the reaction of the alkoxy radical with the NO2 radical (Reaction 1.10) may be important for formation of the C1 – C2 alkyl nitrates (Flocke et al., 1998; Simpson et al., 2006; Archibald et al., 2007). Archibald et al. (2007) reported a model run which suggested that the formation of methyl nitrate via the two production routes would be equal at NOx values of 37 ppb. It is unclear from the paper whether they mean the NOx mixing ratio or NO2 mixing ratio but either would suggest a value of a similar magnitude to that shown in Table 6.6.]

\subsubsection*{Loss reactions}
Being a reservoir for the short lived NOx and having atmospheric lifetimes ranging from several days to about a month \citep{Reeves2007} alkyl nitrates can be transported over long distances, and then lost through photolysis and reaction with $OH$:
\begin{equation} \label{reac:RONO2+hv=RO+NO2}
RONO_2 + hv \rightarrow RO + NO_2
\end{equation}
\begin{equation} \label{reac:RONO2+OH=RCHO+NO2}
RONO_2 + OH \rightarrow RCHO + NO_2
\end{equation}
It means that alkyl nitrates can ‘release’ $NO_2$ back to the atmosphere after a while, and in association with reactions (10) and (11) start a new $O_3$ formation cycle being quite away from the source.
\citep{Newland2013}
[The main loss processes for the alkyl nitrates are oxidation by the OH radical and photodissociation (e.g. Talukdar et al., 1997a, 1997b). The importance of the OH sink relative to photolysis increases with increasing carbon chain length to become the dominant loss process in C5 alkyl nitrates and above (Roberts and Fajer, 1989).]

The alkyl nitrates are removed from the atmosphere through photolysis and reaction with hydroxyl radical.
The relative importance of these two loss reactions depends generally on the net UV intensity and the size of the carbon chain in alkyl nitrates. Photolysis is relatively more important for nitrates with $< C_4$, while nitrates with longer chains ($\geq C_5$) tend to react faster with $OH$. In case of butyl nitrates both mechanisms are of similar significance \citep{Worton2010}.

Alkyl nitrates are removed from the atmosphere by photolysis \citep{Turberg1990} or through oxidation by hydroxyl radical ($OH$) \citep{Talukdar1997}. Photolysis is the dominant loss mechanism for the compounds with short carbon chains and becomes more important with decreasing carbon number. Reaction with $OH$ follows the opposite trend and is most important for the pentyl nitrates and higher.

\subsection{Ozone, alkyl nitrates and carbon bonds}
The question that is always important for ozone-policies is the ozone production efficiency. Historically people noticed that with decreasing NOx emission the ozone production? concentration? drops, so the way of calculating the efficiency of ozone production due to drop in NOx emissions was proposed in the following way:

@@

\citep{deKock1994}
Air samples were collected at Cape Recife (34 ° 1' S 25 ° 40'E) situated on the south-east coast of Southern Afi'ica (Figure 1). The suburban, urban and industrial areas of Port Elizabeth lie approximately 10-40 kin north to north-west of the sampling site.
Mean sumANs 12.7.
Variations in absolute and relative concentrations of the alkyl nitrates could be related to changes in meteorological conditions. It is considered that the differences observed in chemical composition may represent different sources of trace gas emissions or different degrees of photochemical processing of similar sources. The meas~ed levels of the alkyl nitrates suggest long-range transport to play a major role in regulating the distribution of alkyl nitrates in the troposphere.

@@

\subsubsection*{Lifetime}
\citep{Newland2013}
[Species within this group provide a transport reservoir for the short lived NOx compounds due to their relatively longer lifetimes. This allows NOx to be transported from polluted areas to remote background sites. Alkyl nitrates have been reported as making up only a few percent of the NOy ('odd' nitrogen) budget in urban and rural areas (Shepson et al., 1993; Thornberry et al., 2001), however due to their relatively long lifetimes and their continued production during transport of polluted air masses, as alkanes are converted to alkyl nitrates (e.g. Bertman et al., 1995), they can comprise up to tens of percent of the NOy budget in polar regions and in the remote marine troposphere (Talbot et al., 2000). Therefore alkyl nitrates could potentially be a substantial source of NOx in remote background regions contributing to tropospheric ozone formation in these regions.
The lifetimes of the alkyl nitrates vary with latitude and season due to the seasonality of the OH sink and photolysis. Clemitshaw et al. (1997) report a range of lifetimes with respect to OH and to photolysis. Photolysis lifetimes vary from about ten days for butyl and pentyl nitrate in the summer to 1 – 2 years in the winter at 65N with lifetimes also decreasing with altitude. The lifetimes due to reaction with OH (tauOH) at 65N range from tens of years in
winter to about 10 days in the summer. tauOH increases with altitude as the OH concentration decreases. At the equator tauOH is similar all year round, on the order of a few days.]

>>>>

\subsection{Alkyl nitrates}
Alkyl nitrates ($RONO_2$) are organic trace gases, which through one of their formation mechanisms are closely linked with the production of tropospheric ozone \citep{Reeves2007}. This mechanism is present in the following section.

\subsubsection*{Alkyl nitrates chemistry}
The photochemical formation of the alkyl nitrates starts when chemistry of the troposphere reaches the regime of $O_3$ production, i.e. when reactions of the peroxy radicals with $NO$ start to compete with their self- and cross-reactions more effectively due to an increase in $NO_x$ mixing ratio. In this case the reaction between the alkyl peroxy radical ($RO_2$, where $R$ denotes methyl group, $CH_3$) and $NO$ can reveal its two-channel nature \citep{Day2003}:
\begin{subequations} \label{reac:RONO2alpha}
\begin{align}
RO_2 + NO &\xrightarrow{1-\alpha_3} RO + NO_2 \label{reac:RO+NO=aphaRO+NO2}\\
RO_2 + NO &\xrightarrow{\alpha_3} RONO_2 \label{reac:RO+NO=aphaRONO2}
\end{align}
\end{subequations}
The latter reaction can serve as a termination step of the propagated $O_3$ forming cycle (Figure ???), and as \citep{Roberts1998} estimated it is likely to begin to occur when the $NO$ mixing ratio increases up to 100-200 ppt. The yield of alkyl nitrate formation is controlled by the branching ratio ($\alpha_3$), which is temperature and pressure dependent and increases with the number of carbons in the alkyl peroxy radical \citep{Roberts1990}.

An alternative photochemical pathway of the formation of alkyl nitrates is the reaction of the alkoxy radical ($RO$) with $NO_2$:
\begin{equation}\label{reac:RO+NO2=RONO2}
RO + NO_2 \rightarrow RONO_2
\end{equation}
This pathway has been shown to become dominant over reaction (\ref{reac:RO+NO=aphaRONO2}) only for the production of lower (methyl ($C_1$), ethyl ($C_2$)) nitrates when $NO_x$ mixing ratio exceeds 35 ppb \citep{Archibald2007}, i.e. in highly polluted environments.

Alkyl nitrates are removed from the atmosphere by photolysis \citep{Turberg1990} and $OH$ radical oxidation \citep{Talukdar1997}:
\begin{equation} \label{reac:RONO2+hv=RO+NO2}
RONO_2 + hv \rightarrow RO + NO_2
\end{equation}
\begin{equation} \label{reac:RONO2+OH=RCHO+NO2}
RONO_2 + OH \rightarrow RCHO + NO_2
\end{equation}
The relative importance of these two loss reactions depends generally on the net UV intensity and the size of the carbon chain in the alkyl nitrate. Photolysis is relatively more important for nitrates with $< C_4$, while nitrates with longer chains ($\geq C_5$) tend to react faster with $OH$ \citep{Worton2010}.

The lifetimes of alkyl nitrates depend on the seasonality of the photolysis and $OH$ sink, which in turn vary with latitude and altitude. According to \citep{Clemitshaw1997} and \citep{Turberg1990} alkyl nitrates have atmospheric lifetimes of several days to several years, which makes them temporary reservoirs for the short lived $NO_x$ compounds. Being transported from polluted areas to remote background regions and eventually destroyed alkyl nitrates can 'release' $NO_2$ back to the atmosphere, thereby contributing to tropospheric ozone formation.
\subsection{Ozone, alkyl nitrates and carbon bonds}
The question that is always important for ozone-policies is the ozone production efficiency. Historically people noticed that with decreasing NOx emission the ozone production? concentration? drops, so the way of calculating the efficiency of ozone production due to drop in NOx emissions was proposed in the following way:

@@

\citep{deKock1994}
Air samples were collected at Cape Recife (34 1' S 25 40'E) situated on the south-east coast of Southern Afi'ica (Figure 1). The suburban, urban and industrial areas of Port Elizabeth lie approximately 10-40 kin north to north-west of the sampling site.
Mean sumANs 12.7.
Variations in absolute and relative concentrations of the alkyl nitrates could be related to changes in meteorological conditions. It is considered that the differences observed in chemical composition may represent different sources of trace gas emissions or different degrees of photochemical processing of similar sources. The meas~ed levels of the alkyl nitrates suggest long-range transport to play a major role in regulating the distribution of alkyl nitrates in the troposphere.

@@
